---
title: "BIOST/EPI 554 Problem Set 2"
author: "Yichen Lu"
date: "2/17/2021"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE, echo = FALSE}
# rm(list = ls())
knitr::opts_chunk$set(echo = FALSE, include=FALSE, message = FALSE, warning = FALSE, cache = TRUE)
library(SpatialEpi)
library(RColorBrewer)
library(ggplot2)
library(ggridges)
library(INLA)
library(rgdal)
library(spdep)
library(dplyr)
library(fields)
library(maptools)
library(geoR)
library(sp)
library(gstat)
options("scipen"=100)
```

#### PROBLEM 1
(a) Examine the level of clustering of the residuals using Moran’s statistic.

```{r}
## PROBLEM 1
# (a)

# summarize and merge data based on hw1 solution
ohio_data <- read.table("C:/Users/luy85/Google Drive (luyichen1994@gmail.com)/UW Seattle/BIOST554/data/ohio_data-1.txt", header = TRUE, sep = " ")
ohio_county <- ohio_data %>% group_by(fips) %>%
  summarise(year = unique(year),
            deaths = sum(deaths),
            popn = sum(popn),
            exp = unique(county_level_E))

ohio_map <- readOGR(dsn = "C:/Users/luy85/Google Drive (luyichen1994@gmail.com)/UW Seattle/BIOST554/data", layer = "ohio_map")
ohio_map@data$deaths <- ohio_county$deaths
ohio_map@data$exp <- ohio_county$exp
ohio_map@data$popn <- ohio_county$popn

tmp1 <- merge(x = ohio_map@data[,c("STATEFP00",
                                   "COUNTYFP00",
                                   "CNTYIDFP00",
                                   "NAME00")],
              y = ohio_county,
              by.x = 'CNTYIDFP00',
              by.y = 'fips',
              sort = FALSE)

tmp2 <- ohio_county[match(ohio_map@data$CNTYIDFP00,
                          ohio_county$fips),]
tmp2$county_name <- ohio_map@data$NAME00[match(tmp2$fips, ohio_map@data$CNTYIDFP00)]
ohio_county <- tmp2
rm(tmp1, tmp2)

# create a neighborhood list
ohio.nb <- poly2nb(ohio_map)

# evaluate moran's test
col.W <- nb2listw(ohio.nb, style = "W", zero.policy = TRUE)
quasipmod <- glm(deaths~1,offset=log(exp),
                 data=ohio_map@data,family=quasipoisson())
sidsres <- residuals(quasipmod,type="pearson")
m.test <- moran.test(sidsres,col.W)
p <- m.test$p.value
```

*Spatial autocorrelation is evaluated using Moran's test with the "W" style weight function. We mapped the residuals to get a visual on the clustering. Based on the p-value from Moran's test `r p`, we reject the null hypothesis that there's clustering at a significance level of 0.05.*

```{r, include = TRUE}
#map the residuals to get a visual on the clustering
ohio_map$res <- sidsres
par(mar=c(.1,.1,.1,.1))
spplot(ohio_map,"res", main=list(label="Map of residuals (no clustering detected)",cex=1.5))
```


(b) Examine the level of clustering of the residuals using Geary’s statistic.

```{r}
g.test <- geary.test(sidsres, col.W)
p <- g.test$p.value
```

*We also evaluated spatial autocorrelation with Geary's test with the "W" style weight function. We Based on the p-value from Geary's test `r p`, we reject the null hypothesis that there's clustering at a significance level of 0.05.*

#### PROBLEM 2
Cluster Detection for the Ohio lung cancer data. Perform cluster detection on these data using the SatScan method of Kulldorff. Use a significance level of 0.05, and a maximum population size of 20%

```{r, include = FALSE}
## problem 2
# geo <- for(i in 1:88){ohio_map@polygons[[i]]@labpt}
# gCentroid
geo <- latlong2grid(coordinates(ohio_map))

# referencep <- sum(ohio_map@data$deaths)/sum(ohio_map@data$popn)
population <- ohio_map@data$popn
cases <- ohio_map@data$deaths
expected <- ohio_map@data$exp

#significance level of 0.05, and a maximum population size of 20%
pop.upper.bound <- 0.2
n.simulations <- 1000
alpha.level <- 0.05

#StatScan
Kpoisson <- kulldorff(geo,cases,population,expected.cases=NULL, 
                      pop.upper.bound, n.simulations, alpha.level, plot=T)
Kcluster <- Kpoisson$most.likely.cluster$location.IDs.included
```

*The location IDs identified as the most likely cluster include `r Kcluster`. They are plotted in the graph below.*

```{r, include = TRUE}
# plot
ohio_poly = SpatialPolygons(ohio_map@polygons)
plot(ohio_poly, axes = TRUE)
plot(ohio_poly[Kcluster], add = TRUE, col = "red")
title("Most Likely Cluster")
```

#### PROBLEM 3
In the `geoR` library there are data `ca20` which we will explore/model using various geostatistical techniques.

(a) Examine cloud and binned semi-variograms and comment on the evidence for spatial dependence.

```{r, include = TRUE}
## Problem 3
# (a)
# variogram cloud
cloudca20 <- variog(ca20, option = "cloud")
plot(cloudca20, ylab="Semi-variance",xlab="Distance",col="grey",
     main = "Variogram cloud, constant mean",
     cex=.4)

# binned semi-variograms
binca20 <- variog(ca20,uvec=seq(0,5000,100),
                  trend=~ca20[[3]]$area+ca20[[3]]$altitude)
plot(binca20,ylab="Semi-variance",xlab="Distance",
     main = "Binned variogram with linear trend in area andd altitude",
     cex=.5,col="blue")
```

*We take area and altitude as predictor variables. The binned semi-variogram was produced with a model that removed the linear effects of area and altitude on calcium content measurements. Based on the binned semi-variogram, we conclude there's spatial correlation.*

(b) Using the semi-variogram examine Monte Carlo intervals for no spatial dependence.

```{r, include = TRUE}
# (b) 
geoca20.env <- variog.mc.env(ca20, obj=binca20)
plot(binca20,env=geoca20.env,
     main = "Monte Carlo envelops under no spatial dependence",
     xlab="Distance",ylab="Semi-variance")
```

*Based on the Monte Carlo envelopes under no spatial dependence, we conclude that there is dependence here.*

(c) Fit an exponential variogram model to these data, using least squares, maximum likelihood and restricted maximum likelihood.

```{r, include = FALSE}
# (c)
# estimate the initial values
phi_est <- -600/log(0.05)
sigma_est <- 100

#OLS
olsfit <- variofit(binca20,ini=c(sigma_est, phi_est), weights = "equal")
olsfit

# WLS
wlsfit <- variofit(binca20,ini=c(sigma_est, phi_est))
wlsfit

# Maximum likelihood
mlfit <- likfit(ca20,ini=c(sigma_est, phi_est),
                trend=~ca20[[3]]$area+ca20[[3]]$altitude)
mlfit$parameters.summary

# Restricted maximum likelihood
remlfit <- likfit(ca20,ini=c(sigma_est, phi_est),lik.method="RML",
                  trend=~ca20[[3]]$area+ca20[[3]]$altitude)
remlfit$parameters.summary
```

*We first give initial estimates for $\sigma^2$ and $\phi$ using the binned variogram: $\tilde{d}$ = 600 to give $\tilde{\phi} = `r phi_est`, $\tilde{\sigma}$ = sigma_est.*

*Then we estimate model parameters using OLS, WLS, maximum likelihood and restricted maximum likelihood.*

- In OLS, we used equal weights and got $\tilde{\tau}^2$ = 41.3137, $\tilde{\sigma}^2$ = 51.6473, and $\tilde{\phi}$ = 105.6286. 
- Using WLS where we weighted by the number of pairs in each bin, we got $\tilde{\tau}^2$ = 35.8052, $\tilde{\sigma}^2$ = 67.3760, and $\tilde{\phi}$ = 148.1355. 
- For MLE, we got $\tilde{\tau}^2$ = 0.0000, $\tilde{\sigma}^2$ = 104.3638, and $\tilde{\phi}$ = 73.0004, which are very different from the LS estimates*
- For RMLE, we got $\tilde{\tau}^2$ = 22.61870, $\tilde{\sigma}^2$ = 137.6734, and $\tilde{\phi}$ = 200.2841*

```{r, include = TRUE}
plot(binca20,max.dist=1200, xlab="Distance (m)",ylab="Semi-variance",pch=19,cex=.6)
lines(olsfit,max.dist=1200,col="red")
lines(wlsfit,max.dist=1200,lty=2,col="green")
lines(mlfit,max.dist=1200,lty=3,col="blue")
lines(remlfit,max.dist=1200,lty=4,col="black")
legend("bottomright",legend=c("OLS","WLS","ML","REML"),
       lty=c(1,2,3,4),bty="n",col=c("red","green","blue","black"),
       cex=0.5)
```

(d) Carry out kriging and examine the resultant surface, both in terms of the mean and the standard deviation.

*We first produced an image plot of the surface of the residuals. To get the estimated mean, we calculate the mean of the observed data points using the coefficients from the ML fit. Then we plot the surface of the mean by adding the predicted residual to the mean value of the observed data points. We see that area close to the observed data points have very small residuals. Lastly. we plot the Kriging standard deviation of the predictions and area close to the observed datapoints have very small SD.*

```{r, include = TRUE}
# (d)
# form a geodata object with the residuals as the response
lmfit <- lm(ca20$data~ca20[[3]]$area+ca20[[3]]$altitude)
detrend <- as.geodata(cbind(ca20$coords,lmfit$residuals))

# MLE on the detrended data
mlfit2 <- likfit(detrend,ini=c(sigma_est, phi_est))

# ordinary kriging
# summary(detrend$coords)
pred.grid <- expand.grid(seq(4900,6000,l=51), seq(4800, 5800, l=51))
kc <- krige.conv(detrend,loc=pred.grid, krige=krige.control(obj.m=mlfit2))

# residual plot
image.plot(x=pred.grid[["Var1"]][1:51],
           y=unique(pred.grid[["Var2"]]),
           z=matrix(kc$predict,nrow=51,ncol=51),col=terrain.colors(100),
           xlab="x-coordinate",ylab="y-coordinate", 
           main = "Spatial surface of residuals from Kriging")
symbols(detrend$coords[,1],detrend$coords[,2],
        circles=(detrend$data-min(detrend$data))/1,add=T,inches=0.04)

# resultant surface of the mean
m <- mean(lmfit$coefficients[1] + 
            lmfit$coefficients[2] * (ca20[[3]]$area == 2) + 
            lmfit$coefficients[3] * (ca20[[3]]$area == 3) + 
            lmfit$coefficients[4] * ca20[[3]]$altitude) # calculate mean(b0 + b1x_i) using current data

image.plot(x=pred.grid[["Var1"]][1:51],
           y=unique(pred.grid[["Var2"]]),
           z=matrix(m + kc$predict,nrow=51,ncol=51),col=terrain.colors(100),
           xlab="x-coordinate",ylab="y-coordinate", 
           main = "Kriging: resultant surface of the mean")
symbols(detrend$coords[,1],detrend$coords[,2],
        circles=(detrend$data-min(detrend$data))/1,add=T,inches=0.04)

# resultant surface of SD
image.plot(x=pred.grid[["Var1"]][1:51],
           y=unique(pred.grid[["Var2"]]),
           z=matrix(sqrt(kc$krige.var),nrow=51,ncol=51),
           col=cm.colors(100),
           xlab="x-coordinate",ylab="y-coordinate",
           main = "Kriging: resultant surface of SD")
points(detrend$coords[,1],detrend$coords[,2],pch=16)
```


#### Code

```{r, eval = FALSE, echo = TRUE, include = TRUE}
## PROBLEM 1
# (a)

# summarize and merge data based on hw1 solution
ohio_data <- read.table("C:/Users/luy85/Google Drive (luyichen1994@gmail.com)/UW Seattle/BIOST554/data/ohio_data-1.txt", header = TRUE, sep = " ")
ohio_county <- ohio_data %>% group_by(fips) %>%
  summarise(year = unique(year),
            deaths = sum(deaths),
            popn = sum(popn),
            exp = unique(county_level_E))

ohio_map <- readOGR(dsn = "C:/Users/luy85/Google Drive (luyichen1994@gmail.com)/UW Seattle/BIOST554/data", layer = "ohio_map")
ohio_map@data$deaths <- ohio_county$deaths
ohio_map@data$exp <- ohio_county$exp
ohio_map@data$popn <- ohio_county$popn

tmp1 <- merge(x = ohio_map@data[,c("STATEFP00",
                                   "COUNTYFP00",
                                   "CNTYIDFP00",
                                   "NAME00")],
              y = ohio_county,
              by.x = 'CNTYIDFP00',
              by.y = 'fips',
              sort = FALSE)

tmp2 <- ohio_county[match(ohio_map@data$CNTYIDFP00,
                          ohio_county$fips),]
tmp2$county_name <- ohio_map@data$NAME00[match(tmp2$fips, ohio_map@data$CNTYIDFP00)]
ohio_county <- tmp2
rm(tmp1, tmp2)

# create a neighborhood list
ohio.nb <- poly2nb(ohio_map)

# evaluate moran's test
col.W <- nb2listw(ohio.nb, style = "W", zero.policy = TRUE)
quasipmod <- glm(deaths~1,offset=log(exp),
                 data=ohio_map@data,family=quasipoisson())
sidsres <- residuals(quasipmod,type="pearson")
m.test <- moran.test(sidsres,col.W)
p <- m.test$p.value

#map the residuals to get a visual on the clustering
ohio_map$res <- sidsres
par(mar=c(.1,.1,.1,.1))
spplot(ohio_map,"res", main=list(label="Map of residuals (no clustering detected)",cex=1.5))

g.test <- geary.test(sidsres, col.W)
p <- g.test$p.value

## problem 2
# geo <- for(i in 1:88){ohio_map@polygons[[i]]@labpt}
# gCentroid
geo <- latlong2grid(coordinates(ohio_map))

# referencep <- sum(ohio_map@data$deaths)/sum(ohio_map@data$popn)
population <- ohio_map@data$popn
cases <- ohio_map@data$deaths
expected <- ohio_map@data$exp

#significance level of 0.05, and a maximum population size of 20%
pop.upper.bound <- 0.2
n.simulations <- 1000
alpha.level <- 0.05

#StatScan
Kpoisson <- kulldorff(geo,cases,population,expected.cases=NULL, 
                      pop.upper.bound, n.simulations, alpha.level, plot=T)
Kcluster <- Kpoisson$most.likely.cluster$location.IDs.included

# plot
ohio_poly = SpatialPolygons(ohio_map@polygons)
plot(ohio_poly, axes = TRUE)
plot(ohio_poly[Kcluster], add = TRUE, col = "red")
title("Most Likely Cluster")

## Problem 3
# (a)
# variogram cloud
cloudca20 <- variog(ca20, option = "cloud")
plot(cloudca20, ylab="Semi-variance",xlab="Distance",col="grey",
     main = "Variogram cloud, constant mean",
     cex=.4)

# binned semi-variograms
binca20 <- variog(ca20,uvec=seq(0,5000,100),
                  trend=~ca20[[3]]$area+ca20[[3]]$altitude)
plot(binca20,ylab="Semi-variance",xlab="Distance",
     main = "Binned variogram with linear trend in area andd altitude",
     cex=.5,col="blue")

# (b) 
geoca20.env <- variog.mc.env(ca20, obj=binca20)
plot(binca20,env=geoca20.env,
     main = "Monte Carlo envelops under no spatial dependence",
     xlab="Distance",ylab="Semi-variance")

# (c)
# estimate the initial values
phi_est <- -600/log(0.05)
sigma_est <- 100

#OLS
olsfit <- variofit(binca20,ini=c(sigma_est, phi_est), weights = "equal")
olsfit

# WLS
wlsfit <- variofit(binca20,ini=c(sigma_est, phi_est))
wlsfit

# Maximum likelihood
mlfit <- likfit(ca20,ini=c(sigma_est, phi_est),
                trend=~ca20[[3]]$area+ca20[[3]]$altitude)
mlfit$parameters.summary

# Restricted maximum likelihood
remlfit <- likfit(ca20,ini=c(sigma_est, phi_est),lik.method="RML",
                  trend=~ca20[[3]]$area+ca20[[3]]$altitude)
remlfit$parameters.summary

plot(binca20,max.dist=1200, xlab="Distance (m)",ylab="Semi-variance",pch=19,cex=.6)
lines(olsfit,max.dist=1200,col="red")
lines(wlsfit,max.dist=1200,lty=2,col="green")
lines(mlfit,max.dist=1200,lty=3,col="blue")
lines(remlfit,max.dist=1200,lty=4,col="black")
legend("bottomright",legend=c("OLS","WLS","ML","REML"),
       lty=c(1,2,3,4),bty="n",col=c("red","green","blue","black"),
       cex=0.5)

# (d)
# form a geodata object with the residuals as the response
lmfit <- lm(ca20$data~ca20[[3]]$area+ca20[[3]]$altitude)
detrend <- as.geodata(cbind(ca20$coords,lmfit$residuals))

# MLE on the detrended data
mlfit2 <- likfit(detrend,ini=c(sigma_est, phi_est))

# ordinary kriging
# summary(detrend$coords)
pred.grid <- expand.grid(seq(4900,6000,l=51), seq(4800, 5800, l=51))
kc <- krige.conv(detrend,loc=pred.grid, krige=krige.control(obj.m=mlfit2))

# residual plot
image.plot(x=pred.grid[["Var1"]][1:51],
           y=unique(pred.grid[["Var2"]]),
           z=matrix(kc$predict,nrow=51,ncol=51),col=terrain.colors(100),
           xlab="x-coordinate",ylab="y-coordinate", 
           main = "Spatial surface of residuals from Kriging")
symbols(detrend$coords[,1],detrend$coords[,2],
        circles=(detrend$data-min(detrend$data))/1,add=T,inches=0.04)

# resultant surface of the mean
m <- mean(lmfit$coefficients[1] + 
            lmfit$coefficients[2] * (ca20[[3]]$area == 2) + 
            lmfit$coefficients[3] * (ca20[[3]]$area == 3) + 
            lmfit$coefficients[4] * ca20[[3]]$altitude) # calculate mean(b0 + b1x_i) using current data

image.plot(x=pred.grid[["Var1"]][1:51],
           y=unique(pred.grid[["Var2"]]),
           z=matrix(m + kc$predict,nrow=51,ncol=51),col=terrain.colors(100),
           xlab="x-coordinate",ylab="y-coordinate", 
           main = "Kriging: resultant surface of the mean")
symbols(detrend$coords[,1],detrend$coords[,2],
        circles=(detrend$data-min(detrend$data))/1,add=T,inches=0.04)

# resultant surface of SD
image.plot(x=pred.grid[["Var1"]][1:51],
           y=unique(pred.grid[["Var2"]]),
           z=matrix(sqrt(kc$krige.var),nrow=51,ncol=51),
           col=cm.colors(100),
           xlab="x-coordinate",ylab="y-coordinate",
           main = "Kriging: resultant surface of SD")
points(detrend$coords[,1],detrend$coords[,2],pch=16)
```
